Linting Linearity in Core/System FC

### 2023-01-22 report.tex
notas

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar
in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
disjunction operator ($\oplus$) and the $\bang$ modality operator. A typing
judgement for the linearly typed lambda calculus has the form
%
\[ \Gamma; \Delta \vdash M : A \]
%
where $\Gamma$ is the context of resources that may be used unrestrictedly, that
is, any number of times, $\Delta$ is the context of resources that must be used
linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
When resources from the linear context are used, they are removed from the
context and no longer available, and all resources in the linear context must be
used exactly once.

@@

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\qquad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \qquad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\qquad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\label{fig:llcrules}
\end{figure}

@@

Haskell is a functional programming language defined by the "Haskell
Report"\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when the value is needed. The combination of purety and laziness is
unique to Haskell among mainstream programming languages...

Haskell is a large language but its small core is based on a typed lambda
calculus. As such, there exist no statements and computation is done simply
through the evaluation of functions. Besides functions, one can define types and
their constructors and pattern match on said constructors. Function application
is denoted by an empty space (\texttt{f~a} means \texttt{f} applied to
\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
by the enumerated alternatives. All variable names start with lower case and
types start with upper case (excluding type variables). To make explicit the
type of an expression, the \texttt{::} operator is used (e.g.
$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
from \texttt{Int} to \texttt{Bool}).

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
non-nullary\footnote{\texttt{IO} has kind $\texttt{Type}\to\texttt{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\texttt{IO~Int}, \texttt{IO~Bool}); \texttt{IO} by itself is a \emph{type
constructor}} type constructor \texttt{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textt{NonEmpty} is a list that can not be empty).

@@

A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \texttt{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
which users can define data types as they would common recursive algebraic data
types, with the added abilitiy to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches. Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

>>>>

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus\todo{citar o girard
  provavelmente ou o Barber (\url{https://www.lfcs.inf.ed.ac.uk/reports/96/ECS-LFCS-96-347/})}, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar
in Fig.~\ref{fig:llcgrammar} and what programs are well typed through the typing
rules in Fig.~\ref{fig:llcrules}. The language features functions and function application ($\lolli$), two
flavours of pairs, additive ($\with$) and multiplicative ($\tensor$), a
disjunction operator ($\oplus$)\todo{sum types} and the $\bang$
modality operator\todo{que serve para}. A typing
judgement for the linearly typed lambda calculus has the form
%
\[ \Gamma; \Delta \vdash M : A \]
%
where $\Gamma$ is the context of resources that may be used unrestrictedly, that
is, any number of times, $\Delta$ is the context of resources that must be used
linearly (\emph{exactly once}), $M$ is the program to type and $A$ is its type.
When resources from the linear context are used, they are removed from the
context and no longer available, and all resources in the linear context must be
used exactly once.

@@

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\quad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \quad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\quad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\label{fig:llcrules}
\end{figure}

@@

Haskell is a functional programming language defined by the Haskell
Report~\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when its value is needed. The combination of purity and laziness is
unique to Haskell among mainstream programming languages.

Haskell is a large feature-rich language but its relatively small core is based on a typed lambda
calculus. As such, there exist no statements and computation is done simply
through the evaluation of functions. Besides functions, one can define types and
their constructors and pattern match on said constructors. Function application
is denoted by an empty space\todo{juxtaposition} (\texttt{f~a} means \texttt{f} applied to
\texttt{a}) and pattern matching is done with the \texttt{case} keyword followed
by the enumerated alternatives. All variable names start with lower case and
types start with upper case (excluding type variables). To make explicit the
type of an expression, the \texttt{::} operator is used (e.g.
$\texttt{f~::~Int}\to \texttt{Bool}$ is read \texttt{f} \emph{has type} function
from \texttt{Int} to \texttt{Bool}).

Because Haskell is a pure programming language, input/ouptut side-effects are
``made pure'' by modelling them on the type level through the
non-nullary\footnote{\texttt{IO} has kind $\texttt{Type}\to\texttt{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\texttt{IO~Int}, \texttt{IO~Bool}); \texttt{IO} by itself is a \emph{type
constructor}} type constructor \texttt{IO}.\todo{IO monad; made pure
nao quer dizer nada. A ideia é que valores do tipo IO t capturam
\emph{computacoes} que quando executadas produzem efeitos e produzem
valores do tipo $t$, os operadores monadicos permitem compor
computacoes, mas nao executam nada, por isso e que e tudo puro.}
Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \texttt{NonEmpty} is a list that can not be empty).

@@

A well liked motto is
"make invalid states unrepresentable". In this light, we will\todo{nao
  uses o futuro} introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \texttt{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
which users can define data types as they would common algebraic data
types, with the added ability to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches.
\todo{nao ha branches, é nos varios construtores. This allows for
  additional type invariants to be represented with GADTs through
  their type parameters, which restricts the use of specific
  constructors and their subsequent deconstruction through pattern
  matching}

Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives.
\todo{acho que a outra frase que escrevi e melhor que esta / complementa}
We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

  \todo[inline]{Acho que a parte dos naturais nao e necessaria}
